```{r}
# Uncomment the following lines if you need to install packages:
# install.packages(c("dplyr", "ggplot2", "stringr"))
# install.packages("arrow")

library(dplyr)
library(ggplot2)
library(stringr)
library(arrow)
```


```{r}
# Define target PTM parameters (similar to a Python tuple)
TARGET_PTM <- c("Citrullination", "UniMod:7", "R")

# Set the path to the PTM data (parquet file)
PTM_DATA <- "/Users/cgu3/Documents/diann/data/RAINBOW_UNIPROT_human_revi_2024_12_19_ProteinAG_citrullination_report-lib.parquet"

```


# Data loading
```{r}
# Read metadata CSV
metadata <- readr::read_csv("data/metadata.csv")
print(metadata)

# Read the parquet file using the arrow package
dat_raw <- read_parquet(PTM_DATA)
print(dat_raw)
```


```{r}
# random sample 1% as test data
dat_sample <- dat_raw[sample(nrow(dat_raw), nrow(dat_raw) * 0.01), ]
# save as parquet
write_parquet(dat_sample, "data/RAINBOW_UNIPROT_human_revi_2024_12_19_ProteinAG_citrullination_report-lib_sample.parquet")
```


```{r}
# Left join metadata with the citrullination data on column "ID"
dat <- dat_raw %>% inner_join(metadata, by = "ID", suffix = c("", ".meta"))
head(dat)
```


# Create PTM Indicator Columns
```{r}
# Check if 'Modified.Sequence' contains TARGET_PTM[2] ("UniMod:7")
dat$has_target_PTM <- grepl(TARGET_PTM[2], dat$Modified.Sequence)

# Check if 'Stripped.Sequence' contains TARGET_PTM[3] ("R")
dat$has_target_site <- grepl(TARGET_PTM[3], dat$Stripped.Sequence)

# Count occurrences of TARGET_PTM[2] in 'Modified.Sequence'
dat$num_PTM <- str_count(dat$Modified.Sequence, fixed(TARGET_PTM[2]))

```


# Aggregate Total Intensity

```{r}
total_intensity <- dat %>%
  group_by(plate, ID, assay, evotip, well, has_target_PTM) %>%
  summarise(total_intensity = sum(`Precursor.Quantity`, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(total_intensity = log10(total_intensity)) %>%
  left_join(metadata, by = "ID", suffix = c("", ".meta")) %>% 
  # rename columns with .meta
  select(-ends_with(".meta")) %>% 
  filter(!is.na(`Cancer Type`))


print(total_intensity)

```

# Plot: Histogram by Group for Each Assay
```{r}

# Aggregate the data by summing intensities grouped by 'group', 'ID', and 'assay'
aggregated_intensity <- total_intensity %>%
  group_by(plate, ID, assay, evotip, well) %>%
  summarise(total_intensity = sum(total_intensity, na.rm = TRUE), .groups='drop') %>% 
  left_join(metadata, by = "ID")


# Define colors for groups (e.g., Case and Control)
case_colors <- c("blue", "orange")

# Loop through unique assays and plot histograms
unique_assays <- unique(aggregated_intensity$assay)
for (ass in unique_assays) {
  subset_assay <- aggregated_intensity %>% filter(assay == ass)
  p <- ggplot(subset_assay, aes(x = total_intensity, fill = group)) +
         geom_histogram(bins = 100, alpha = 0.5, position = "identity") +
         labs(title = paste("Distribution of total intensity by Case/Control for assay", ass),
              x = "Total intensity (log10)",
              y = "Frequency") +
         theme_minimal()
  print(p)
}

```

# Plot: Histogram by Combination of Group and Modification Status
```{r}
# Create a combined label for group and PTM status
total_intensity_labeled <- total_intensity_labeled %>%
  mutate(ptm_label = ifelse(has_target_PTM, "Modified peptide", "Unmodified peptide"),
         combined = paste(group, ptm_label, sep = ", "))


p <- ggplot(total_intensity_labeled, aes(x = total_intensity, fill = combined)) +
  geom_histogram(bins = 100, alpha = 0.5, position = "identity") +
  labs(x = "Total intensity (log10)", y = "Frequency",
       title = paste("Distribution of total intensity of peptide with", TARGET_PTM[1],
                     "on site", TARGET_PTM[3])) +
  theme_minimal()
print(p)

```

# Plot: Histogram by Cancer Type for Assay "IgB"
```{r}
total_intensity_IgB <- total_intensity %>%
  filter(has_target_PTM, assay == "IgB", !is.na(`Cancer Type`))

p <- ggplot(total_intensity_IgB, aes(x = total_intensity, fill = `Cancer Type`)) +
  geom_histogram(bins = 30, alpha = 0.3, position = "identity") +
  labs(x = "Total intensity (log10)", y = "Frequency",
       title = paste("Distribution of total intensity of peptide with", TARGET_PTM[1],
                     "on site", TARGET_PTM[3], "by Cancer Type")) +
  theme_minimal()
print(p)

```
# Aggregation at Peptide and Protein Levels
```{r}
# Peptide-level aggregation: group by ID, assay, and Stripped.Sequence; take any TRUE for has_target_PTM
peptide_level_aggregation <- dat %>%
  filter(has_target_site == TRUE) %>%
  group_by(plate, ID, assay, evotip, well, Stripped.Sequence) %>%
  summarise(has_target_PTM = any(has_target_PTM), 
            total_intensity = sum(Precursor.Quantity), .groups='drop') %>% 
  left_join(metadata, by = "ID")

# Protein-level aggregation: group by ID, assay, and Protein.Group; take any TRUE for has_target_PTM
protein_level_aggregation <- dat %>%
  filter(has_target_site == TRUE) %>%
  group_by(plate, ID, assay, evotip, well, Protein.Group) %>%
  summarise(has_target_PTM = any(has_target_PTM),
            total_intensity = sum(Precursor.Quantity), .groups = 'drop') %>% 
  left_join(metadata, by = "ID")


# Summarise the mean of has_target_PTM for each assay at the peptide level
peptide_level_summary <- peptide_level_aggregation %>%
  group_by(assay) %>%
  summarise(mean_has_target_PTM = mean(has_target_PTM))
print(peptide_level_summary)

# Summarise the mean of has_target_PTM for each assay at the protein level
protein_level_summary <- protein_level_aggregation %>%
  group_by(assay) %>%
  summarise(mean_has_target_PTM = mean(has_target_PTM))
print(protein_level_summary)

```

# Plot: Distribution of Proportion of Modified Proteins

```{r}


# Calculate the mean of has_target_PTM for each ID, assay, and group
mean_has_PTM <- protein_level_aggregation %>%
  group_by(ID, assay, group) %>%
  summarise(mean_has_target_PTM = mean(has_target_PTM)) %>%
  ungroup()

unique_assays <- unique(mean_has_PTM$assay)
for (ass in unique_assays) {
  subset_assay <- mean_has_PTM %>% filter(assay == ass)
  p <- ggplot(subset_assay, aes(x = mean_has_target_PTM, fill = group)) +
         geom_histogram(bins = 50, alpha = 0.5, position = "identity") +
         labs(x = "Proportion of modified peptides", y = "Frequency",
              title = paste("Distribution of proportion of modified protein with", TARGET_PTM[1],
                            "on site", TARGET_PTM[3], "for assay", ass)) +
         scale_fill_manual(values = c("blue", "orange")) +
         theme_minimal()
  print(p)
}

```

# Plot: Histogram of Total Number of Peptides with PTM (Peptide-Level Aggregation)
```{r}

# Sum up the number of peptides with PTM (TRUE counts as 1, FALSE as 0) for each ID, group, and assay
total_peptides_with_PTM <- peptide_level_aggregation %>%
  group_by(ID, group, assay) %>%
  summarise(total_peptides_with_PTM = sum(as.numeric(has_target_PTM))) %>%
  ungroup()

unique_assays <- unique(total_peptides_with_PTM$assay)
for (ass in unique_assays) {
  subset_assay <- total_peptides_with_PTM %>% filter(assay == ass)
  p <- ggplot(subset_assay, aes(x = total_peptides_with_PTM, fill = group)) +
         geom_histogram(bins = 50, alpha = 0.5, position = "identity") +
         labs(x = "Total number of peptides with PTM", y = "Frequency",
              title = paste("Histogram of Total Number of Peptides with", TARGET_PTM[1],
                            "on site", TARGET_PTM[3], "by group for assay", ass)) +
         scale_fill_manual(values = c("blue", "orange")) +
         theme_minimal()
  print(p)
}

```



```{r}
# Create violin plota
total_intensity %>% 
  ggplot(aes(x = factor(plate), y = `total_intensity`)) +
  geom_violin(trim = FALSE, fill = "skyblue", alpha = 0.5) +  # Violin plot with fill color
  geom_boxplot(width = 0.1, outlier.shape = NA) +  # Boxplot inside violin for median and quartiles
  facet_wrap(~assay + has_target_PTM, labeller = labeller(has_target_PTM = c("TRUE" = "Modified peptides", "FALSE" = "Unmodified peptides"))) +
  theme_minimal() +  # Minimal theme for cleaner look
  labs(x = "Batch ID", y = "log10(Precursor.Quantity)", title = "Total intensity of samples group by plate") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels if needed

```
```{r}
modified_peptide_level <- dat %>%
  group_by(plate, ID, assay, evotip, well, Modified.Sequence) %>%
  summarise(Genes=first(Genes),
            Protein.Group=first(Protein.Group),
            Protein.Ids=first(Protein.Ids),
    has_target_PTM = any(has_target_PTM), 
            total_intensity = sum(Precursor.Quantity), .groups='drop')
```



```{r}
# pivot the data into wide format with modified.Sequence as rows and ID in column
modified_peptide_level %>% 
  # mutate(total_intensity = log10(total_intensity)) %>%
  pivot_wider(id_cols=c(Protein.Group, Protein.Ids, Genes, Modified.Sequence), names_from = c(plate, ID, assay, evotip, well), values_from = total_intensity)
```

```{r}
# this is weird to have the same peptide with different protein.Group
modified_peptide_level %>% filter(ID=='MCED_1' &  assay=='FT' & Modified.Sequence=='AAC(UniMod:4)LLPKLDELRDEGK')
```

```{r}
modified_peptide_level  %>% select(plate, ID, assay, evotip, well) %>% unique() %>% mutate(Run=paste(plate, ID, assay, evotip, well, sep='_')) %>% inner_join(metadata, by='ID')
```

